---
title: "Mixed Effect Modeling"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(ggplot2)
library(dplyr)
library(lmerTest)
library(pander)
library(broom)
library(kableExtra)
library(car)
```

# Load Data
```{r}
# Regular data
homelessness_data = read.csv('../data/processed/pivoted_and_PIT.csv')

#Standardized Data
standardize = function(col){
  col = (col - mean(col)) / sd(col)
  return(col)
}

homelessness_data_standard = homelessness_data |>
  mutate(across(where(is.numeric) & !all_of("Unsheltered.Per.100.000"), standardize))
```

# Why Mixed Effects Model
```{r}
# Examine variation in Unsheltered.Per.100.000 by City
ggplot(homelessness_data, aes(x = City, y = Unsheltered.Per.100.000)) +
  geom_boxplot() +
  labs(title = "Distribution of Unsheltered Rate per 100,000 by City") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
This graph shows variability based on city, and if we want to focus on effects of programs on unsheltered rather than city influences on those programs, it would make sense to treat city as a random effect

# Implement Models
```{r}
ignored_features = c("PEH.Per.100.000", "Total.PEH", "Unsheltered.PEH", "Population")

model <- lmer(Unsheltered.Per.100.000 ~
              bridge.to.housing.network + emergency.shelter + family.reunification.program +
              flexible.funds + food.and.nutrition + homeless.services + homelessness.prevention +
              homeshare.program + housing.assistance + housing.navigation.services +
              housing.stability.services + motel.voucher + neighborhood.revitalization.services +
              opening.doors.program + outreach + project.h.o.p.e. + rapid.re.housing +
              rental.assistance + restrooms + safe.parking + service.center + staff.and.operations +
              take.back.the.streets + transitional.housing + work.for.hope +
              (1  | City), data = homelessness_data |> select(-all_of(ignored_features)))

#Standarized model
modelS <- lmer(Unsheltered.Per.100.000 ~
              bridge.to.housing.network + emergency.shelter + family.reunification.program +
              flexible.funds + food.and.nutrition + homeless.services + homelessness.prevention +
              homeshare.program + housing.assistance + housing.navigation.services +
              housing.stability.services + motel.voucher + neighborhood.revitalization.services +
              opening.doors.program + outreach + project.h.o.p.e. + rapid.re.housing +
              rental.assistance + restrooms + safe.parking + service.center + staff.and.operations +
              take.back.the.streets + transitional.housing + work.for.hope +
              (1 | City), data = homelessness_data_standard |> select(-all_of(ignored_features)))

```
# Addressing Warnings
```{r}
vif(model)
```
The VIF for work.for.hope is extremely large (4.08e+13) with an infinite value for the scaled GVIF, indicating severe multicollinearity, likely because this variable is almost perfectly collinear with one or more other predictors in the model. This is the source of the "rank deficient" warning, and R removes work.for.hope from the model.

# Visualizing City-Level Random Effects to Justify Mixed-Effects Model Usage
```{r}
# Random effects for City
ranef_city <- ranef(model)$City
ggplot(ranef_city, aes(x = reorder(row.names(ranef_city), `(Intercept)`), y = `(Intercept)`)) +
  geom_point() +
  coord_flip() +
  labs(title = "City-Level Random Effects", x = "City", y = "Random Intercept Estimate")

```


# Model Summaries
```{r, results='asis'}
# For PDF
summarize_model <- function(model) {
  # Extract the full model summary
  model_summary <- summary(model)

  # Convert the fixed effects table to a data frame
  fixed_effects <- as.data.frame(model_summary$coefficients)

  # Sort the fixed effects by Estimate
  sorted_fixed_effects <- fixed_effects[order(fixed_effects$Estimate), ]

  # Add the C-like format for estimates, but remove the predictor names in the formatted column
  sorted_fixed_effects$Estimates <- sprintf("%.4f", sorted_fixed_effects$Estimate)

  # Remove the original 'Estimate' column
  sorted_fixed_effects <- sorted_fixed_effects[, !names(sorted_fixed_effects) %in% c("Estimate")]

  # Rearrange the columns to move 'Estimates' to the front
  sorted_fixed_effects <- sorted_fixed_effects[, c("Estimates", setdiff(names(sorted_fixed_effects), "Estimates"))]

  # Print the modified summary with formatted estimates
  sorted_fixed_effects %>%
    kable("latex", booktabs = TRUE) %>%
    kable_styling(latex_options = "scale_down")
}

# For Console
# summarize_model <- function(model) {
#   # Extract the full model summary
#   model_summary <- summary(model)
#   
#   # Convert the fixed effects table to a data frame
#   fixed_effects <- as.data.frame(model_summary$coefficients)
#   
#   # Sort the fixed effects by Estimate
#   sorted_fixed_effects <- fixed_effects[order(fixed_effects$Estimate), ]
#   
#   # Add the C-like format for estimates, but remove the predictor names in the formatted column
#   sorted_fixed_effects$Estimates <- sprintf("%.4f", sorted_fixed_effects$Estimate)
#   
#   # Remove the original 'Estimate' column
#   sorted_fixed_effects <- sorted_fixed_effects[, !names(sorted_fixed_effects) %in% c("Estimate")]
#   
#   # Rearrange the columns to move 'Estimates' to the front
#   sorted_fixed_effects <- sorted_fixed_effects[, c("Estimates", setdiff(names(sorted_fixed_effects), "Estimates"))]
#   
#   # Print the modified summary with formatted estimates
#   print(sorted_fixed_effects)
# }

summarize_model(model)
summarize_model(modelS)
```
```{r, results='asis', echo=FALSE}
#for pdf formatting
cat("\\newpage")
```

# Plot Estimates
```{r}
# Extract fixed effects
fixed_effects <- as.data.frame(summary(model)$coefficients)

# Remove the intercept row
fixed_effects_no_intercept <- fixed_effects[rownames(fixed_effects) != "(Intercept)", ]

# Reorder the predictors based on the Estimate
fixed_effects_no_intercept$Predictor <- factor(
  rownames(fixed_effects_no_intercept),
  levels = rownames(fixed_effects_no_intercept)[order(fixed_effects_no_intercept$Estimate)]
)

# Plot the fixed effects excluding the intercept, sorted by Estimate
ggplot(fixed_effects_no_intercept, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 10), # Rotate and adjust the labels
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  ) +
  labs(
    title = "Fixed Effects Coefficients",
    x = "Predictors",
    y = "Estimate"
  )

# Extract fixed effects
fixed_effectsS <- as.data.frame(summary(modelS)$coefficients)

# Remove the intercept row
fixed_effects_no_interceptS <- fixed_effectsS[rownames(fixed_effectsS) != "(Intercept)", ]

# Reorder the predictors based on the Estimate
fixed_effects_no_interceptS$Predictor <- factor(
  rownames(fixed_effects_no_interceptS),
  levels = rownames(fixed_effects_no_interceptS)[order(fixed_effects_no_interceptS$Estimate)]
)

# Plot the fixed effects excluding the intercept, sorted by Estimate
ggplot(fixed_effects_no_interceptS, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 10), # Rotate and adjust the labels
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  ) +
  labs(
    title = "Fixed Effects Coefficients For Standarzied Model",
    x = "Predictors",
    y = "Estimate"
  )

```
Flexible funds in both graphs as one of lower 3, and housing stability services in both graphs as upper 3. Lower indicates decrease in unsheltered as funds allocated for program, and upper indicates increase.

# Assessing Residuals and Model Fit
```{r}
# Plot residuals
plot(resid(model))
qqnorm(resid(model))
qqline(resid(model))
```
Variance in residuals in constant for most part, and no overall patterns which is a good sign. There are some outliers which can be examined

# Results if outliers are removed
```{r, warning=FALSE}
# Extract residuals from the model
residuals <- resid(model)

# Set a threshold for identifying outliers
# For example, residuals greater than 2 standard deviations from the mean
threshold <- 2 * sd(residuals)

# Identify the indices of the outliers
outlier_indices <- which(abs(residuals) > threshold)

# Display the indices and residual values of the outliers
outliers <- data.frame(Index = outlier_indices, Residual = residuals[outlier_indices])
print(outliers)

# View the corresponding rows in the original data
outlier_data <- homelessness_data[outlier_indices, ]

# Remove the identified outlier indices from the dataset
homelessness_data_no_outliers <- homelessness_data[-c(15, 17, 32, 33), ]

# Re-run the model without outliers
model_no_outliers <- lmer(Unsheltered.Per.100.000 ~ 
              bridge.to.housing.network + emergency.shelter + family.reunification.program + 
              flexible.funds + food.and.nutrition + homeless.services + homelessness.prevention + 
              homeshare.program + housing.assistance + housing.navigation.services + 
              housing.stability.services + motel.voucher + neighborhood.revitalization.services + 
              opening.doors.program + outreach + project.h.o.p.e. + rapid.re.housing + 
              rental.assistance + restrooms + safe.parking + service.center + staff.and.operations + 
              take.back.the.streets + transitional.housing + work.for.hope + 
              (1 | City), 
              data = homelessness_data_no_outliers |> select(-all_of(ignored_features)))

summarize_model(model_no_outliers)

#Standardized Model with no outliers
homelessness_data_no_outliers_standard = homelessness_data_no_outliers |>
  mutate(across(where(is.numeric) & !all_of("Unsheltered.Per.100.000"), standardize))

model_no_outliersS <- lmer(Unsheltered.Per.100.000 ~ 
              bridge.to.housing.network + emergency.shelter + family.reunification.program + 
              flexible.funds + food.and.nutrition + homeless.services + homelessness.prevention + 
              homeshare.program + housing.assistance + housing.navigation.services + 
              housing.stability.services + motel.voucher + neighborhood.revitalization.services + 
              opening.doors.program + outreach + project.h.o.p.e. + rapid.re.housing + 
              rental.assistance + restrooms + safe.parking + service.center + staff.and.operations + 
              take.back.the.streets + transitional.housing + work.for.hope + 
              (1 | City), 
              data = homelessness_data_no_outliers_standard |> select(-all_of(ignored_features)))

summarize_model(model_no_outliersS)

#Graph estmates for models without ouliers
# Extract fixed effects
fixed_effects <- as.data.frame(summary(model)$coefficients)

# Remove the intercept row
fixed_effects_no_intercept <- fixed_effects[rownames(fixed_effects) != "(Intercept)", ]

# Reorder the predictors based on the Estimate
fixed_effects_no_intercept$Predictor <- factor(
  rownames(fixed_effects_no_intercept),
  levels = rownames(fixed_effects_no_intercept)[order(fixed_effects_no_intercept$Estimate)]
)

# Plot the fixed effects excluding the intercept, sorted by Estimate
ggplot(fixed_effects_no_intercept, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 10), # Rotate and adjust the labels
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  ) +
  labs(
    title = "Fixed Effects Coefficients Without Outliers",
    x = "Predictors",
    y = "Estimate"
  )

# Extract fixed effects
fixed_effectsS <- as.data.frame(summary(model_no_outliersS)$coefficients)

# Remove the intercept row
fixed_effects_no_interceptS <- fixed_effectsS[rownames(fixed_effectsS) != "(Intercept)", ]

# Reorder the predictors based on the Estimate
fixed_effects_no_interceptS$Predictor <- factor(
  rownames(fixed_effects_no_interceptS),
  levels = rownames(fixed_effects_no_interceptS)[order(fixed_effects_no_interceptS$Estimate)]
)

# Plot the fixed effects excluding the intercept, sorted by Estimate
ggplot(fixed_effects_no_interceptS, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 10), # Rotate and adjust the labels
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  ) +
  labs(
    title = "Fixed Effects Coefficients For Standarzied Model Without Outliers",
    x = "Predictors",
    y = "Estimate"
  )

# Residuals and Normality without outliers
plot(resid(model_no_outliers))
qqnorm(resid(model_no_outliers))
qqline(resid(model_no_outliers))

```
The order of most to least efficient programs remains consistent even when outliers are removed, only the extent to which they affect Unsheltered PEH per 100,000 is affected, though not by much.

